Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-16 paper.tex
edits to mutation appendix

First, mutations are placed on the tree sequence under an inhomogeneous Poisson model
by applying them independently to each edge.
If an edge spans a region $[a, b)$ of the genome
and connected parent and child nodes with times $s < t$,
and the mutation rate locally is $\mu$,
then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
and each mutation is placed independently at a position chosen uniformly in $[a,b)$
and a time uniformly in $[s,t)$.
In a discrete genome, all positions are integers and so more than one mutation
may occur at the same position on the same edge.
Otherwise (i.e., for an infinite-sites model),
positions are rejection sampled to obtain a unique floating-point number.
If an edge spans a region of the genome with more than one mutation rate,
this is done separately for each distinct region.
The algorithm scales with the number of edges in the tree sequence.
Since the number of edges is proportional to the number of marginal trees,
an algorithm that applies mutations separately to each marginal tree
would take at least as many steps,
and would require additional bookkeeping necessary to place mutations uniformly on a tree.

Next, alleles are chosen for each mutation.
If the site was not previously mutated, then a new ancestral allele is chosen for the site,
according to an input distribution of ancestral state allele probabilities.
Then, each mutation on the tree is considered in turn,
in order so that older mutations are considered first,
and a derived allele is chosen based on the parental allele
(which may be the ancestral allele or the derived allele of a previous mutation)
and some randomness.
Finally, information about the mutations are recorded in the site and mutation tables
of the tree sequence.

@@

Since the algorithm above applies mutations at a single rate
independent of ancestral state,
a model in which different alleles mutate at different rates
must necessarily produce some silent mutations,
i.e., mutations in which the derived allele is equal to the parental allele.
To illustrate this, consider a mutation model on the binary alphabet
$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
To implement this, first place mutations at the largest total rate, which is $\alpha$.
Then, at each site, choose an ancestral allele from the ``root distribution''
(a natural choice would be to choose from the stationary distribution,
$X$ with probability $\beta / (\alpha + \beta)$),
and for each mutation, choose a derived allele as follows:
if the parental allele is $X$, then choose $Y$;
if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
and choose $Y$ otherwise.
This produces the correct distribution by Poisson thinning:
a Poisson process with rate $\alpha$ in which each point is discarded independently
with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
process with rate $\beta$.
All finite-state models (implemented under the generic \texttt{MatrixMutationModel} class)
work in this way: mutations are placed at the maximum mutation rate,
and then some silent mutations will result.

In previous versions of \msprime, silent mutations were disallowed,
and we could have removed them from the output entirely.
However, we have chosen to leave them in, so that for instance
simulating with the \texttt{HKY} mutation model will result in silent mutations
if not all equilibrium frequencies are the same.
Although the presence of silent mutations may at first be surprising
(although they don't affect calculations of statistics in any way),
there is a good reason to leave them in:
to allow layering of different mutation models.
Suppose that we wanted to model the mutation process as a mixture of more than one model,
e.g., Jukes-Cantor mutations at rate $\mu_1$, and
HKY mutations occur at rate $\mu_2$.
Layering multiple calls to \texttt{sim\_mutations} is allowed,
so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
and then add more with the \texttt{HKY} model at rate $\mu_2$.
However, there is a small statistical problem:
suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
but then the HKY mutations inserts another mutation in the middle,
resulting in $A \to C \to C$.
If neither mutation model allows silent transitions,
then this is clearly not correct,
i.e., it is not equivalent to a model that simultaneously applies the two models.
(The impact is small, however, as it only affects sites with more than one mutation.)
The solution is to make the Jukes-Cantor model \emph{state-independent}
(also called ``parent-independent''),
by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
\emph{independently} of the parent (so that 1/4 of mutations will be silent).
If so -- and, more generally, if the first mutational process put down is state-independent --
then the result of sequentially applying the two mutation models
is equivalent to the simultaneous model.
To facilitate this, many mutation models have a \texttt{state\_independent} option
that increases the number of silent mutations
and makes the model closer to state-independent.

>>>>

First, mutations are placed on the tree sequence under an inhomogeneous Poisson model
by applying them independently to each edge.
If an edge spans a region $[a, b)$ of the genome
and connected parent and child nodes with times $s < t$,
and the mutation rate locally is $\mu$,
then the number of mutations on the edge is Poisson with mean $\mu (t-s) (b-a)$,
and each mutation is placed independently at a position chosen uniformly in $[a,b)$
and a time uniformly in $[s,t)$.
In a discrete genome, all positions are integers and so more than one mutation
may occur at the same position on the same edge.
Otherwise (i.e., for an infinite-sites model),
positions are rejection sampled to obtain a unique floating-point number.
If an edge spans a region of the genome with more than one mutation rate,
this is done separately for each distinct region.
The algorithm scales with the number of edges in the tree sequence.

Since the number of edges is proportional to the number of marginal trees,
this algorithm is $O(T)$,
where $T$ is the number of trees.
An alternative algorithm would apply mutations separately to each tree.
Since the number of branches per tree is proportional to the number of leaves,
the simplest implementation would be $O(T S)$,
where $S$ is the number of samples.
Any more efficient tree-by-tree algorithm would have to make use of the fact
that structure is shared between adjacent trees---effectively, therefore,
working with shared edges.

Next, alleles are chosen for each mutation.
If the site was not previously mutated, then a new ancestral allele is chosen for the site,
according to an input distribution of ancestral state allele probabilities.
Then, each mutation on the tree is considered in turn,
and a derived allele is randomly chosen based on the parental allele
(which may be the ancestral allele or the derived allele of a previous mutation).
Finally, information about the mutations are recorded in the site and mutation tables
of the tree sequence.

@@

The distribution of ancestral alleles
is used to choose the allele present at the root of the tree at each mutated site,
i.e., the \texttt{root\_distribution}.
Mutation models with a finite possible set of alleles have a natural choice
for this distribution---the \emph{stationary distribution} of the mutation process.
(All mutation models are Markovian,
so this may be found as the top left eigenvector of the mutation matrix.)
This is the default in most models,
except, e.g., the \texttt{BinaryMutationModel}, whose alleles are 0 and 1
and always labels the ancestral allele ``0''.
However, mutational processes are not in general stationary,
so we often allow different root distribution to be specified.

Since the general algorithm above applies mutations at a single rate
independent of ancestral state,
a model in which different alleles mutate at different rates
must necessarily produce some silent mutations,
i.e., mutations in which the derived allele is equal to the parental allele.
To illustrate this, consider a mutation model on the binary alphabet
$\{X, Y\}$ in which $X$ mutates to $Y$ at rate $\alpha$
and $Y$ mutates to $X$ at rate $\beta$, with $\beta < \alpha$.
To implement this, first place mutations at the largest total rate, which is $\alpha$.
Then, at each site, choose an ancestral allele from the ``root distribution''
(a natural choice would be to choose from the stationary distribution,
$X$ with probability $\beta / (\alpha + \beta)$),
and for each mutation, choose a derived allele as follows:
if the parental allele is $X$, then choose $Y$;
if the parental allele is $Y$, then choose $X$ with probability $\beta/(\alpha + \beta)$,
and choose $Y$ otherwise.
This produces the correct distribution by Poisson thinning:
a Poisson process with rate $\alpha$ in which each point is discarded independently
with probability $\beta / (\alpha + \beta)$ is equivalent to a Poisson
process with rate $\beta$.
All finite-state models (implemented under the generic \texttt{MatrixMutationModel} class)
work in this way: mutations are placed at the maximum mutation rate,
% I think I understand what you want to say here, but it would be clearer if an example would have been given - similar to the next section.
and then some silent mutations will result.

In previous versions of \msprime, silent mutations were disallowed,
and we could have removed them from the output entirely.
However, we have chosen to leave them in, so that for instance
simulating with the \texttt{HKY} mutation model will result in silent mutations
if not all equilibrium frequencies are the same.
The presence of silent mutations may at first be surprising
(although they don't affect calculations of statistics in any way),
but there is a good reason to leave them in:
to allow layering of different mutation models.
Suppose that we wanted to model the mutation process as a mixture of more than one model,
e.g., Jukes-Cantor mutations at rate $\mu_1$, and
HKY mutations occur at rate $\mu_2$.
Layering multiple calls to \texttt{sim\_mutations} is allowed,
so we could first apply mutations with the \texttt{JC69} model at rate $\mu_1$
and then add more with the \texttt{HKY} model at rate $\mu_2$.
However, there is a small statistical problem:
suppose that after applying Jukes-Cantor mutations we have an $A \to C$ mutation,
but then the HKY mutations inserts another mutation in the middle,
resulting in $A \to C \to C$.
If neither mutation model allows silent transitions,
then this is clearly not correct,
i.e., it is not equivalent to a model that simultaneously applies the two models.
(The impact is small, however, as it only affects sites with more than one mutation.)
The solution is to make the Jukes-Cantor model \emph{state-independent}
(also called ``parent-independent''),
by placing mutations at rate $4/3 \mu_1$ and then choosing the derived state for each mutation
\emph{independently} of the parent (so that 1/4 of mutations will be silent).
If so -- and, more generally, if the first mutational process put down is state-independent --
then the result of sequentially applying the two mutation models
is equivalent to the simultaneous model.
To facilitate this, many mutation models have a \texttt{state\_independent} option
that increases the number of silent mutations
and makes the model closer to state-independent.
